# Dimensionality reduction techniques

### Overview of the data

*The dataset is a combination of two datasets; [Human Development](http://hdr.undp.org/en/content/human-development-index-hdi) and [Gender Inequality](http://hdr.undp.org/sites/default/files/hdr2015_technical_notes.pdf). After some "data wrangling", the dataset looks like [this](http://s3.amazonaws.com/assets.datacamp.com/production/course_2218/datasets/human2.txt). The wrangling steps can be seen from [Rscript-file in my GitHub](https://github.com/peempe/IODS-project/blob/master/data/create_human.R).*

```{r}
str(human) # 155 observations and 8 variables
summary(human)
```

Let's investigate the association between variables with a correlation matrix:
```{r}
library("dplyr")
library("corrplot")
library("GGally")
ggpairs(human) # Great, an error again. Let's fix it...
ggpairs(human, cardinality_threshold = 155) # Increasing the 'cardinality_threshold' parameter
cor(human) %>% corrplot()
```


### Perform principal component analysis (PCA)

```{r}
pca_human <- prcomp(human) # PCA
biplot(pca_human, choices = 1:2, cex = c(0.8, 1), col = c("grey40", "deeppink2")) #Biplot
```











