# Logistic regression

### Reading the data

```{r results = 'hide'}
alc <- read.table("alc.csv", sep = ";", header = T)
View(alc)
colnames(alc)
```

*The dataset is jointed from two different datasets and it includes information about students' alcohol consumpiton. The infromation is gathered from Portuguese and Math classes from two Portuguese schools*

*More information about the dataset can be found [here](https://archive.ics.uci.edu/ml/datasets/Student+Performance).*


### The research plan and hypotheses

The aim of this week is to analyze the relationship of high/low alcohol consumption (AC) and following variables: mother's education (Medu), quality of family realtionships (famrel), current health status (health), and number of school absences (absences).

**My hypotheses are:**
1. Mother's low education is associated with higher AC.
2. Poor quality of family relationships is associated with higher AC.
3. Poor health status is associated with higher AC.
4. Higher number of school absences is associated with higher AC.


### Preliminary analyses

```{r}
library(ggplot2)
```

**Number of school absences**
```{r}
g1 <- ggplot(alc, aes(x = high_use, y = absences, col = sex))
g1 + geom_boxplot() + ggtitle("School absences by alcohol consumption")
```



